### 2024-06-18 basic-statistics.qmd
20240617 - missing cite
20240617 - formulas

@@

- standard deviation
- variance
- range
- minimum and maximum

@@

@@

Effect size thresholds for small, medium, and large effect sizes are in @tbl-effectSizeThresholds [@McGrath2006].

>>>>

```{r}
library("petersenlab")
library("tidyverse")
```

Descriptive statistics are used to describe data.
For instance, they may be used to describe the center, spread, or shape of the data.
There are various indices of each.

### Center {#sec-descriptiveStatisticsCenter}

Indices to describe the *center* (central tendency) of a variable's data include:

The mean of $X$ ($\bar{X}$) is calculated as:

$$
\bar{X} = \frac{\sum X_i}{n} = \frac{x_1 + x_2 + ... + x_n}{n}
$$ {#eq-IQR}

That is, to compute the mean, sum all of the values and divide by the number of values ($n$).

The median is determined as the value at the 50th percentile (i.e., the value that is higher than 50% of the values and is lower than the other 50% of values).

The mode is the most common value.

Below is R code to estimate each:

@@

- standard deviation
- variance
- range
- minimum and maximum
- interquartile range (IQR)

The (sample) variance of $X$ ($s^2$) is calculated as:

$$
s^2 = \frac{\sum (X_i - \bar{X})^2}{n-1}
$$ {#eq-variance}

The (sample) standard deviation of $X$ ($s$) is calculated as:

$$
s = \sqrt{\frac{\sum (X_i - \bar{X})^2}{n-1}}
$$ {#eq-sd}

The range is calculated of $X$ is calculated as:

$$
\text{range} = \text{maximum} - \text{minimum}
$$ {#eq-range}

The interquartile range (IQR) is calculated as:

$$
\text{IQR} = Q_3 - Q_1
$$ {#eq-IQR}

where $Q_3$ is the score at the third quartile (i.e., 75th percentile), and $Q_1$ is the score at the first quartile (i.e., 25th percentile).

Below is R code to estimate each:

### Shape {#sec-descriptiveStatisticsShape}

@@

Below is R code to estimate each:

### Combination {#sec-descriptiveStatisticsCombination}

To estimate a combination of indices of center, spread, and shape of the data, you can use the following code:

```{r}
#psych::describe(mydata)

#mydata %>% 
#  summarise(across(
#      everything(),
#      .fns = list(
#        n = ~ length(na.omit(.)),
#        missingness = ~ mean(is.na(.)) * 100,
#        M = ~ mean(., na.rm = TRUE),
#        SD = ~ sd(., na.rm = TRUE),
#        min = ~ min(., na.rm = TRUE),
#        max = ~ max(., na.rm = TRUE),
#        range = ~ max(., na.rm = TRUE) - min(., na.rm = TRUE),
#        IQR = ~ IQR(., na.rm = TRUE),
#        median = ~ median(., na.rm = TRUE),
#        mode = ~ petersenlab::Mode(., multipleModes = "mean"),
#        skewness = ~ psych::skew(., na.rm = TRUE),
#        kurtosis = ~ psych::kurtosi(., na.rm = TRUE)),
#      .names = "{.col}.{.fn}")) %>%
#    pivot_longer(
#      cols = everything(),
#      names_to = c("variable","index"),
#      names_sep = "\\.") %>% 
#    pivot_wider(
#      names_from = index,
#      values_from = value)
```

@@

Effect size thresholds [@McGrath2006] for small, medium, and large effect sizes are in @tbl-effectSizeThresholds [@McGrath2006].
